### 2020-03-10 Writing/disparity_review.tex
Updated reference gender balance

\noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids or molluscs), species-rich but phenotypically conservative (e.g. rodents or nematodes), species-poor but phenotypically diverse (e.g. afrotherian mammals), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not necessarily linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve?
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are comparatively more recent. 
Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (e.g. taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace? \citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
    More recently, this approach as also been used to understand the relationship between developmental processes and morphology in the evo-devo field.
    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing researchers to successfully link variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.

	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?

	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. 
	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
	\citep[e.g.][]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.

	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
	\citep{Friedman2010-ve, Pierce2008-yr, Anderson2013-zt}.
	The links between form and function, however, are not always clear cut.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005-or}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{Brusatte2008-vx} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010-ve}.
	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005-or}.

@@

Fundamental insights into evolutionary biology have been elicited from these four types of disparity analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g.][]{Foote1989-fd, Deline2018-le};
	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Anderson2001-qb}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[e.g.][]{Cooney2017-ly}, Fourier coefficients \citep[e.g.][]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep[e.g.][]{Raup1961-vx,Saunders2004} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.

@@

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017-cf}.
This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.

Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Brazeau2017-kg, Palci2018-ni}.
Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset difficult.
Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
Therefore, trait data should be collected with the scientific question in mind, or the question asked should be tailored to the limits of the data available.

@@

Disparity analyses often use ordination techniques for dimension reduction.
Ordinations are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
They come in many flavours depending on the data and the desired morphospace properties.
For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012-va}.

One of the reasons why ordination techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at a time, which can be more intuitive than through disparity indices (see section \ref{section:metrics} below).
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017-ly}.

Like most other aspects of disparity analyses, however, reducing dimensionality can be tricky.
In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results. 
Although a common technique is to consider the \textit{n} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.

Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data is non-Euclidean (e.g. some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
Although some gradients are easy to detect or interpret (e.g. the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[e.g.][]{Weisbecker2019-kp}.
For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0, has no biological meaning if these are binary alternate states.

Categorical data are a good deal more problematic, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e. the distance between A and B is not equal to the distance between B and A][]{Gerber2014-ol}.
Last, but not least, in many cases, ordination might not be necessary.
For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    \tiny{
    Morphospaces: different mathematical representations of a morphospace.
    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}.
    Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
    Visualisation: different  ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation); or ordination axis plots (directly from the ordinated matrix).
    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).
    }
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}

@@

Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
This is usually achieved with a disparity measure or index \citep{Hopkins2017-cf}.
As with any summary of multidimensional data, disparity indices will reflect only some aspects of the morphological variation, never its whole complexity \citep{GuillermeMOMS}.
It is therefore often beneficial to use more than one index to summarise different aspects of variation, guided by the aim of the study.

When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g. volume
\citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g. does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015-qi} - see section \ref{section:ordination} above).

In addition to considering what properties of disparity these indices should capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
For example, measuring the sum of variance-covariance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.

Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}.
Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992-zs} or convex hull based indices \citep{Jackson2011-kq}.

@@

No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018-mg}, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance based on a distance matrix between observations and therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

@@

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account.
It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first $n$ axis of the ordination, and can lead to incorrect interpretations of the data \citep[such as wrongly supporting ``early burst'' type patterns;][]{Uyeda2015}.

@@

\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}.
Innovations in morphological disparity analyses likely await discovery in their respective literatures.

>>>>

\noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids or molluscs), species-rich but phenotypically conservative (e.g. rodents or nematodes), species-poor but phenotypically diverse (e.g. afrotherian mammals), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not necessarily linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve?
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are comparatively more recent. 
Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (e.g. taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995, Briggs1992}.
	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace? \citep{Foote1995, Raup1961, Gerber2017}.
    More recently, this approach as also been used to understand the relationship between developmental processes and morphology in the evo-devo field.
    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing researchers to successfully link variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.

	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010}?

	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. 
	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
	\citep[e.g.][]{Fortey1996, Ruta2013, Hopkins2013}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011}.

	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
	\citep{Pierce2008, Friedman2010}.
	The links between form and function, however, are not always clear cut.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005}.

@@

Fundamental insights into evolutionary biology have been elicited from these four types of disparity analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997, Erwin2007, Hughes2013}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007, Hughes2013}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Friedman2010}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[e.g.][]{Cooney2017}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.

@@

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008,Close2015}.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018}.

Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Palci2018}.
Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset difficult.
Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
Therefore, trait data should be collected with the scientific question in mind, or the question asked should be tailored to the limits of the data available.

@@

Disparity analyses often use ordination techniques for dimension reduction.
Ordinations are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
They come in many flavours depending on the data and the desired morphospace properties.
For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.

One of the reasons why ordination techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at a time, which can be more intuitive than through disparity indices (see section \ref{section:metrics} below).
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017}.

Like most other aspects of disparity analyses, however, reducing dimensionality can be tricky.
In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results. 
Although a common technique is to consider the \textit{n} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018, Gerber2017}.

Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data is non-Euclidean (e.g. some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
Although some gradients are easy to detect or interpret (e.g. the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not \citep[e.g.][]{Weisbecker2019}.
For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0, has no biological meaning if these are binary alternate states.

Categorical data are a good deal more problematic, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e. the distance between A and B is not equal to the distance between B and A][]{Gerber2014}.
Last, but not least, in many cases, ordination might not be necessary.
For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015}.
For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    \tiny{
    Morphospaces: different mathematical representations of a morphospace.
    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015} or an ordinated matrix \citep[e.g. in][]{tyler2011detecting}.
    Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
    Visualisation: different  ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation); or ordination axis plots (directly from the ordinated matrix).
    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).
    }
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}

@@

Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
This is usually achieved with a disparity measure or index \citep{Hopkins2017}.
As with any summary of multidimensional data, disparity indices will reflect only some aspects of the morphological variation, never its whole complexity \citep{GuillermeMOMS}.
It is therefore often beneficial to use more than one index to summarise different aspects of variation, guided by the aim of the study.

When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g. volume
\citealt{Diaz2016}, distance from the centroid \citealt{Hopkins2017, Finlay2015-ft}, variance and range \citealt{Brusatte2008}).
Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g. does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015} - see section \ref{section:ordination} above).

In addition to considering what properties of disparity these indices should capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001, Ciampaglio2001}.
For example, measuring the sum of variance-covariance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012}.

Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966}.
In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966}.
Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992} or convex hull based indices \citep{Jackson2011}.

@@

No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018}, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance based on a distance matrix between observations and therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

@@

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998}.
Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account.
It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first $n$ axis of the ordination, and can lead to incorrect interpretations of the data \citep[such as wrongly supporting ``early burst'' type patterns;][]{Uyeda2015}.

@@

\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013, Saupe2015, Canter2018, mammola2019}.
Innovations in morphological disparity analyses likely await discovery in their respective literatures.

While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
For example, the multidimensional analysis of \citep{Diaz2016}, which analysed patterns of form and function in plants, is essentially an eco-morphospace; isotopic analyses of organisms \citep{Jackson2011,Swanson2015} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013} as an ecosystem-space, etc.
These generalisations could also be exported for any set of traits: cognate approaches have been adopted in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018} where interpretation of the resulting transcriptome-spaces would be improved by giving careful attention to the concerns we highlight concerning morphospaces.

Although disparity analyses are now simple to implement in freely available software \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
We assert that future morphological analyses will benefit from emphasising the methodological decisions made, rather than simply using disparity analysis because it exists.
